Incremental Forward Reasoning for White-Box~Proof~Search

### 2025-02-12 paper.tex
Explain powerset notation

\title{Incremental Forward Reasoning for White-Box Proof Search}

@@

\authorrunning{J.~Limperg and X.~Généreux}
\institute{University of Munich (LMU), Munich, Germany\\
\email{\{jannis.limperg,xavier.genereux\}@lmu.de}}

@@

\begin{abstract}
  Many proof assistants provide automation tactics based on simple, tableau-style tree search, such as Isabelle's and Rocq's auto or Lean's Aesop.
  These tactics solve a given goal by recursively applying \emph{rules}, which return zero or more subgoals.
  In most cases, the rules themselves can be arbitrary tactics, so they can return any set of subgoals corresponding to an admissible inference in the prover's logic.

  In this setting, we consider \emph{forward rules}, which apply a given theorem, say $A → B → C$, to any goal containing hypotheses $A$ and $B$, adding $C$ as an additional hypothesis.
  When treated naively, such rules are tried on every goal encountered during the search, leading to repeated unifications of premises $A$ and $B$ with the hypotheses of each goal.
  Much of this work is redundant since goals often have similar hypotheses.

  We present an implementation of forward rules that takes advantage of these similarities.
  For each goal, we cache certain partial applications of forward rules in a custom data structure, the \emph{partial match state}.
  This enables efficient updates, so if goals have similar hypotheses, we can quickly update the partial match state rather than re-running all forward rules.

  Our technique is compatible with any search strategy and most logics.
  It has been implemented in the Aesop proof search tactic for Lean.

@@

@@

We associate to each goal in Aesop's search tree a partial match state that reflects the goal's context.
Let $\States$ be the set of forward match states, $\Hyps$ the set of hypotheses, $\Rules$ the set of rules and $\CMatches$ the set of complete matches.
The forward match state's interface then consists of two functions:
\begin{align*}
  \addHyp &: \States → \Hyps → \States × \powerset(\Rules × \CMatches) \\
  \delHyp &: \States → \Hyps → \States
\end{align*}
The $\addHyp$ function adds a hypothesis to the partial match state.
Hence, if $s$ is the partial match state corresponding to context $Γ$ and $\addHyp(s, h : T) = (s', M)$, then $s'$ is the partial match state corresponding to $Γ,\, h : T$.
Similarly, if $s$ is the partial match state corresponding to $Γ, h : T$, then $\delHyp(s, h : T)$ is the partial match state corresponding to $Γ$.
The second output of $\addHyp$, $M$, is a set of pairs $(r, m)$ where $r$ is a rule and $m$ is a complete match for $r$.
These are the matches that were completed by adding $h : T$.

>>>>

\title{Incremental Forward Reasoning for White-Box~Proof~Search}

@@

\authorrunning{J.~Limperg and X.~Généreux}
\institute{Ludwig-Maximilians-Universität München, Munich, Germany\\
\email{\{jannis.limperg,xavier.genereux\}@lmu.de}}

@@

\begin{abstract}
  Several proof assistants provide automation tactics based on tableau-style tree search, such as Isabelle's and Rocq's auto and Lean's Aesop.
  We consider \emph{forward rules}, which apply a given theorem, say, $A → B → C$, to any goal containing hypotheses $A$ and $B$, adding $C$ as a hypothesis.
  When treated naively, such rules are tried on every goal encountered during the search, leading to repeated unifications of premises $A$ and $B$ with the hypotheses of each goal.
  We present an approach to forward rules that takes advantage of similarities between successive goals.
  For each goal, we cache partial applications of forward rules in a custom data structure, the \emph{partial match state}, which enables efficient updates.
  Our technique is compatible with any search strategy and most logics.
  It has been implemented in Aesop.

@@

% \begin{abstract}
%   Many proof assistants provide automation tactics based on simple, tableau-style tree search, such as Isabelle's and Rocq's auto or Lean's Aesop.
%   These tactics solve a given goal by recursively applying \emph{rules}, which return zero or more subgoals.
%   In most cases, the rules themselves can be arbitrary tactics, so they can return any set of subgoals corresponding to an admissible inference in the prover's logic.

%   In this setting, we consider \emph{forward rules}, which apply a given theorem, say $A → B → C$, to any goal containing hypotheses $A$ and $B$, adding $C$ as an additional hypothesis.
%   When treated naively, such rules are tried on every goal encountered during the search, leading to repeated unifications of premises $A$ and $B$ with the hypotheses of each goal.
%   Much of this work is redundant since goals often have similar hypotheses.

%   We present an implementation of forward rules that takes advantage of these similarities.
%   For each goal, we cache certain partial applications of forward rules in a custom data structure, the \emph{partial match state}.
%   This enables efficient updates, so if goals have similar hypotheses, we can quickly update the partial match state rather than re-running all forward rules.

%   Our technique is compatible with any search strategy and most logics.
%   It has been implemented in the Aesop proof search tactic for Lean.

%   \keywords{Forward Reasoning \and Forward Chaining \and Tactics \and Interactive Theorem Proving.}
% \end{abstract}

@@

We associate to each goal in Aesop's search tree a partial match state that reflects the goal's context.
Let $\States$ be the set of partial match states, $\Hyps$ the set of hypotheses, $\Rules$ the set of rules, $\CMatches$ the set of complete matches and $\powerset(X)$ the powerset of $X$.
The partial match state's interface then consists of two functions:
\begin{align*}
  \addHyp &: \States → \Hyps → \States × \powerset(\Rules × \CMatches) \\
  \delHyp &: \States → \Hyps → \States
\end{align*}
The $\addHyp$ function adds a hypothesis to the partial match state.
Hence, if $s$ is the partial match state corresponding to context $Γ$ and $\addHyp(s, h : T) = (s', M)$, then $s'$ is the partial match state corresponding to $Γ,\, h : T$.
Similarly, if $s$ is the partial match state corresponding to $Γ, h : T$, then $\delHyp(s, h : T)$ is the partial match state corresponding to $Γ$.
The second output of $\addHyp$, $M$, is a set of pairs $(r, m)$ where $r$ is a rule and $m$ is a complete match for $r$.
These are the matches that were completed by adding $h : T$.
